---
title: "volunteer_clean"
author: "Caroline Gheen"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

=============================================================================
 Project: Manos Juntos STATCOM - Volunteer Check In/Out
 Script: volunteer_clean.R
 Author: Caroline Gheen
 Date: 2026-04-14
 Purpose: Clean and prepare dataset for format in master dataset
=============================================================================

## 1. Packages
## 2. Call in Data and Formatting
## 3. Fix DOB
## 4. Explore Data
## 5. Merge Datasets
## 6. Examine Duplicates
## 8. Duration - Total Time Calculation
## 9. Unique ID Assignment


## 1. Packages
```{r}
# install packages
# install.packages("readxl")
# install.packages("dplyr")
# install.packages("lubridate", repos = "<https://cloud.r-project.org>")
# install.packages("writexl")
```

# Call Packages
```{r}
library(readxl) 
library(dplyr) 
library(lubridate) 
library(stringr)
library(writexl)
library(fuzzyjoin)
library(stringdist)
```


## 2. Call in Data and Formatting
# Call in Data and Format
Read in data, telling each column desired structure
```{r}
data_in <- read_excel(
  'Volunteer_Checkin_Data.xlsx', sheet = 'Sign-In',
  col_types = c(
    "date",    # Timestamp_in
    "text",    # Email Address
    "text",    # First Name
    "text",    # Last Name
    "date",    # Date of Birth
    "text",    # Cell Phone Number
    "text",    # Email (non-school...)
    "skip",    # verification question
    "skip",    # covid
    "skip"     # mask
  )
)
nrow(data_in)
```
6 DOBs were outside expected range - default changed these to be NA

Changes names to be in wanted format
```{r}
data_in <- data_in %>%
  rename(
    timestamp_in   = `Timestamp`,
    email       = `Email Address`,
    first_name  = `First Name`,
    last_name   = `Last Name`,
    dob         = `Date of Birth`,
    phone       = `Cell Phone Number`,
    email_personal   = `Email (non-school email please)`
  )
```

Trim and standardize all free text (first/last name, email)
```{r}
data_in <- data_in %>% 
  mutate( 
    first_name = str_to_lower(str_trim(first_name)), 
    last_name = str_to_lower(str_trim(last_name)), 
    email = str_to_lower(str_trim(email)), 
    email_personal = str_to_lower(str_trim(email_personal)) 
)
```

Read in check out data, telling each column desired structure
```{r}
data_out <- read_excel(
  'Volunteer_Checkin_Data.xlsx', sheet = 'Sign-Out',
  col_types = c(
    "date",    # Timestamp
    "text",    # Email Address
    "text",    # First Name
    "text",    # Last Name
    "date",    # Date of Birth
    "text",    # Cell Phone Number
    "skip",    # verification question
    "skip",    # optional
    "text"     # department
  )
)
nrow(data_out)
```
3 DOB were out of range - default changed to NA


Changes names to be in wanted format out data
```{r}
data_out <- data_out %>%
  rename(
    timestamp_out   = `Timestamp`,
    email       = `Email Address`,
    first_name  = `First Name`,
    last_name   = `Last Name`,
    dob         = `Date of Birth`,
    phone       = `Cell Phone Number`,
    department   = `Which department did you spend the majority of your time volunteering in today?`
  )
```

Trim and standardize all free text (first/last name, email)
```{r}
data_out <- data_out %>% 
  mutate(first_name = str_to_lower(str_trim(first_name)), 
  last_name = str_to_lower(str_trim(last_name)), 
  email = str_to_lower(str_trim(email))
)
```


Load data
```{r}
database <- read_excel("Volunteer_Database_2024.xlsx", col_types = c(
  "numeric", # ID
  "text", # First Name
  "text", # Last Name
  "text", # Volunteer Address 1
  "text", # Volunteer Address 2
  "text", # City
  "text", # State
  "text", # Zip Code
  "text", # Cell
  "text" # Email
))
```

Changes names to be in wanted format
```{r}
database <- database %>% 
    rename( 
    ID = 'ID', 
    first_name = 'First Name',
    last_name ='Last Name', 
    address1 = 'VolunteerAddress1', 
    address2 ='Volunteer Address2', 
    city = 'VolunteerCity',
    state = 'State', 
    zipcode = 'Zip Code', 
    phone = 'Cell', 
    email = 'Volunteer e-mail',
)
```

Trim and standardize all free text (first/last name, email)
```{r}
database <- database %>% 
  mutate( 
    first_name =str_to_lower(str_trim(first_name)), 
    last_name = str_to_lower(str_trim(last_name)), 
    email = str_to_lower(str_trim(email)),
    address1 = str_to_lower(str_trim(address1)), 
    address2 = str_to_lower(str_trim(address2)), 
    city = str_to_lower(str_trim(city)),
    state = str_to_lower(str_trim(state)) 
)
```

Probably should add leading 0s? look up good practice for id numbers

# 3. Fix DOB
Some DOB were misstyped to dates that are impossible (ex: year is 0200 instead of 2000). We can manually fix these DOBs

Manually imput DOB
```{r}
data_in %>% slice(40, 941, 1298, 1380, 1738, 2306) %>% 
  select(first_name, last_name, dob) # data in mistakes

data_out %>% slice(16, 1280, 1281) %>% 
  select(first_name, last_name, dob) # data out mistakes

# manually fix in dob
data_in <- data_in %>%
  mutate(dob = case_when(
    first_name == "cade" & last_name == "huston" ~ as.Date("1998-09-14"),
    first_name == "paul" & last_name == "atiyeh" ~ as.Date("2004-06-01"),
    first_name == "athman" & last_name == "rajaratnam" ~ as.Date("2001-02-20"),
    first_name == "emily" & last_name == "nguyen" ~ as.Date("2000-09-13"),
    first_name == "adrian" & last_name == "hernandez" ~ as.Date("2003-03-03"),
    first_name == "megan" & last_name == "john" ~ as.Date("2003-10-21"),
    TRUE ~ dob
  ))

# manually fix dob out
data_out <- data_out %>%
  mutate(dob = case_when(
    first_name == "cade" & last_name == "huston" ~ as.Date("1998-09-14"),
    first_name == "emily" & last_name == "nguyen" ~ as.Date("2000-09-13"),
    first_name == "lena" & last_name == "do" ~ as.Date("1998-11-08"),
    TRUE ~ dob
  ))

# check all dob present
sum(is.na(data_in$dob)) #0
sum(is.na(data_out$dob)) #0
```

# 4. Explore Data
Explore data (sign-in and out
```{r}
dim(data_in) #3805 and 7 variables
dim(data_out) #3701 and 7 variables - 104 people did not check out
```

Show unique values of each to determine how many different people used check in/out, presumably - not perfect due to typos
```{r}
nrow(unique(data_in[c("first_name","last_name")])) #463 people
nrow(unique(data_out[c("first_name","last_name")])) #457 people
```

Need to extract the date out of the timestamps so that they can match - if you try and match on timestamp outright there will be no matches as timestamp is very specific, down to the second
```{r}
data_in <- data_in %>% mutate(date = as.Date(timestamp_in))
data_out <- data_out %>% mutate(date = as.Date(timestamp_out))
```

Is anyone missing time in/time out
```{r}
sum(is.na(data_out$timestamp_out)) # 0
sum(is.na(data_in$timestamp_in)) # 0
```

# 5. Merge Data (first name, last name, dob is best)
Used left join so that people who checked in but not out, still remained in the merged dataset

Goal is for nrow(data_in) == nrow(merged). using data_in as basis so matching all instances; if there are extra rows, it is due to a many-many issue of the matching not being specific enough and R creating all possible combinations (Ex: john and sara both born on same date volunteer same date, R make 4 possible combinations)
*nrow(data_in) = 3805*
*look at accidental duplication due to system*

I wanted to try two ways to compare approaches, but only 19 people have an email in sign in... is this true??
```{r}
sum(!is.na(data_in$email)) # 19 people do have an email
sum(is.na(data_in$dob))  # 6 people do not have a dob - default
```

Merge based on same email and date
```{r}
merged_email <- data_in %>% 
left_join(data_out, by = c("email", "date"))
nrow(merged_email)
```

Merge based on same first, last and date
```{r}
merged_name <- data_in %>% 
left_join(data_out, by = c("first_name","last_name", "date")) 
nrow(merged_name) # 3821
```

Merge based on same dob, first, last and date
*BEST SO FAR*
```{r}
merged_name_dob <- data_in %>% 
left_join(data_out, by = c("first_name", "last_name", "date", "dob")) 
nrow(merged_name_dob) #3818 -  13 extra, best so far

test <- inner_join(data_in, data_out, by = c("first_name", "last_name", "date", "dob"))

test2 <- anti_join(data_in, data_out, by = c("first_name", "last_name", "date", "dob"))

nrow(test) #3397
nrow(test2) #421
nrow(merged_name_dob) - nrow(test) #421
```
*Only 18 many-to-many issues. Looking at the inner join (rows that exist in both so all entries) there is 421 extra meaning 421 in data_out got dropped due to either no check in or could not find a match.*
*All not so honest because there are people who entered multiple times and typos...*

Merge based on same dob, and date
```{r}
merged_dob <- data_in %>% 
left_join(data_out, by = c("date", "dob"))
nrow(merged_dob) # 3894
```

When we merge based on first and last name, there is a warning that some first+last+date combination appears multiple times. It could be people accidentally filling out the check in when they meant to scan the check out; people doing two shifts in one day; or people with the same name. Most likely the first one, but we shouldn't assume until we confirm
*Common Reasons to Have Duplicates on Check In: check in twice; check in instead of out; check in and did not check out*
*Common Reasons to Have Duplicates on Check Out: people thought they were checking out and stayed longer so they checked out twice; checked out instead of in at the beginning; forgot to check in but checked out*

# Duplicates 
starting at merged_name_dob chosen and named as merged
```{r, warning = FALSE}
nrow(data_in) # 3805
nrow(data_out) # 3701
nrow(data_in) - nrow(data_out) ## 104

merged <- data_in %>% 
left_join(data_out, by = c("date", "dob", "first_name", "last_name"))
nrow(merged) # 3818
```

assign row id individually to data_in and data_out
```{r}
data_in  <- data_in  %>% mutate(id_in  = row_number())
data_out <- data_out %>% mutate(id_out = row_number())
```

Fix duplicate check in - people who submitted two check ins instead of check in check out will have the second check in occurance go to check out
*44 people had duplicate check ins for same day*
```{r}
# how many people had more than one checkin matching on first, last, date, dob
duplicates_in <- data_in %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) %>%
  ungroup()

nrow(duplicates_in)  # 44

unique_dupes_in <- duplicates_in %>%
  distinct(first_name, last_name, date, dob)

for (i in 1:nrow(unique_dupes_in)) {
  fn    <- unique_dupes_in$first_name[i]
  ln    <- unique_dupes_in$last_name[i]
  dt    <- unique_dupes_in$date[i]
  dob_i <- unique_dupes_in$dob[i]  

  # check if they have a checkout already
  has_checkout <- nrow(
    data_out %>% filter(first_name == fn, last_name == ln, date == dt, dob == dob_i)
  ) > 0

  if (has_checkout) {
    # has a checkout - delete all but first check-in
    rows_to_remove <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%  # keep first, remove the rest
      pull(id_in)

    data_in <- data_in %>% filter(!id_in %in% rows_to_remove)

  } else {
    # no checkout - move all extra check-ins to data_out
    extra_rows <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%  # everything except the first
      rename(timestamp_out = timestamp_in) %>%
      mutate(department = NA_character_) %>%
      select(-email_personal)

    data_out <- bind_rows(data_out, extra_rows)

    # delete the extra check-ins from data_in
    rows_to_remove <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%
      pull(id_in)

    data_in <- data_in %>% filter(!id_in %in% rows_to_remove)
  }
}

# verify - should return 0 rows
data_in %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1)
```

Fix duplicate check out - people who submitted two check outs instead of check in check out will have the first check out occurance go to check in
*36 people had duplicate check outs for same day*
```{r}
# how many people had more than one check out matching on first, last, date, dob
duplicates_out <- data_out %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) %>%
  ungroup()

nrow(duplicates_out)  #36

unique_dupes_out <- duplicates_out %>%
  distinct(first_name, last_name, date, dob)

for (i in 1:nrow(unique_dupes_out)) {
  fn    <- unique_dupes_out$first_name[i]
  ln    <- unique_dupes_out$last_name[i]
  dt    <- unique_dupes_out$date[i]
  dob_i <- unique_dupes_out$dob[i]

  # check if they have a check-in already
  has_checkin <- nrow(
    data_in %>% filter(first_name == fn, last_name == ln, date == dt, dob == dob_i)
  ) > 0

  if (has_checkin) {
    # has a check-in - delete all but first checkout
    rows_to_remove <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%  # keep first, remove the rest
      pull(id_out)

    data_out <- data_out %>% filter(!id_out %in% rows_to_remove)

  } else {
    # no check-in - move first checkout to data_in
    extra_row <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(1) %>%
      rename(timestamp_in = timestamp_out) %>%
      mutate(email_personal = NA_character_) %>%
      select(-department)

    data_in <- bind_rows(data_in, extra_row)

    # delete all duplicates from data_out
    rows_to_remove <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      pull(id_out)

    data_out <- data_out %>% filter(!id_out %in% rows_to_remove)
  }
}

# verify - should return 0 rows
data_out %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1)
```

```{r}
# 1. did row counts change as expected?
nrow(data_in)# 3788
nrow(data_out) #3692

# 2. are there any remaining duplicates in data_in?
data_in %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) #0

# 3. are there any remaining duplicates in data_out?
data_out %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) #0

# remerge
merged <- data_in %>% 
left_join(data_out, by = c("date", "dob", "first_name", "last_name"))
nrow(merged) # 3788
nrow(data_in) #3788
```
# Create Duration
Ok, now there are no extra check in occurances. Make duration column and check how many NA
```{r}
# calculate duration
merged <- merged %>%
  mutate(duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")))

# then check NAs
sum(is.na(merged$duration)) #398 NA

# and see what they look like
merged %>%
  filter(is.na(duration)) %>%
  select(first_name, last_name, date, dob, timestamp_in, timestamp_out)
```

Clean id columns from rbind
```{r}
# remove artifact columns from data_in and data_out
data_in  <- data_in  %>% select(-any_of("id_out"))
data_out <- data_out %>% select(-any_of("id_in"))

# reassign clean IDs
data_in  <- data_in  %>% mutate(id_in  = row_number())
data_out <- data_out %>% mutate(id_out = row_number())

# remerge cleanly
merged <- data_in %>%
  left_join(data_out, by = c("date", "dob", "first_name", "last_name"))

names(merged)  # should now be clean with no .x .y on IDs
```

Pull in all unmatched checkouts
```{r}
# these are unmatched AND clean enough to have joined correctly
unmatched_out_clean <- data_out %>%
  anti_join(merged, by = c("first_name", "last_name", "date", "dob"))
nrow(unmatched_out_clean) #302

# but the TRUE unmatched pool is actually the NAs in merged
truly_unmatched_in <- merged %>%
  filter(is.na(timestamp_out))
nrow(truly_unmatched_in) #398
```
# Create Notes Column 
Clean data in of extra wording in either first name in or out and append to note column 
```{r}
# initialize note column in data_in
if (!"note" %in% names(data_in)) {
  data_in <- data_in %>% mutate(note = NA_character_)
}

data_in <- data_in %>%
  mutate(
    new_note_first = case_when(
      str_detect(first_name, "[ /]") ~ str_extract(first_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note_last = case_when(
      str_detect(last_name, "[ /]") ~ str_extract(last_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note = case_when(
      !is.na(new_note_first) & !is.na(new_note_last) ~ paste(new_note_first, new_note_last, sep = " | "),
      !is.na(new_note_first) ~ new_note_first,
      !is.na(new_note_last)  ~ new_note_last,
      TRUE ~ NA_character_
    ),
    note = case_when(
      is.na(note) & is.na(new_note) ~ NA_character_,
      is.na(note)                   ~ new_note,
      is.na(new_note)               ~ note,
      TRUE                          ~ paste(note, new_note, sep = " | ")
    ),
    first_name = str_trim(str_extract(first_name, "^[^ /]+")),
    last_name  = str_trim(str_extract(last_name,  "^[^ /]+"))
  ) %>%
  select(-new_note_first, -new_note_last, -new_note)
```

Clean data out of extra wording in either first name in or out and append to note column 
```{r}
# initialize note column in data_out
if (!"note" %in% names(data_out)) {
  data_out <- data_out %>% mutate(note = NA_character_)
}

data_out <- data_out %>%
  mutate(
    new_note_first = case_when(
      str_detect(first_name, "[ /]") ~ str_extract(first_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note_last = case_when(
      str_detect(last_name, "[ /]") ~ str_extract(last_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note = case_when(
      !is.na(new_note_first) & !is.na(new_note_last) ~ paste(new_note_first, new_note_last, sep = " | "),
      !is.na(new_note_first) ~ new_note_first,
      !is.na(new_note_last)  ~ new_note_last,
      TRUE ~ NA_character_
    ),
    note = case_when(
      is.na(note) & is.na(new_note) ~ NA_character_,
      is.na(note)                   ~ new_note,
      is.na(new_note)               ~ note,
      TRUE                          ~ paste(note, new_note, sep = " | ")
    ),
    first_name = str_trim(str_extract(first_name, "^[^ /]+")),
    last_name  = str_trim(str_extract(last_name,  "^[^ /]+"))
  ) %>%
  select(-new_note_first, -new_note_last, -new_note)
```

Remerge and reconcile notes 
```{r}
# remerge with clean names
merged <- data_in %>%
  left_join(data_out, by = c("date", "dob", "first_name", "last_name"))

# reconcile notes from both sides into a single note column
merged <- merged %>%
  mutate(
    note = case_when(
      is.na(note.x) & is.na(note.y) ~ NA_character_,
      is.na(note.x)                  ~ note.y,
      is.na(note.y)                  ~ note.x,
      TRUE                           ~ paste(note.x, note.y, sep = " | ")
    )
  ) %>%
  select(-note.x, -note.y)

sum(!is.na(merged$note)) # 37
# verify row count still matches
nrow(merged)   # should equal nrow(data_in) = 3788
```

Recalculate duration and check NA
*15 people had extra writing in their first or last name boxes - this function does split genuine correctly spelled but has a space names as well.*
*37 people have note entries now*
*Future direction: individually audit 37 notes; determine strategy for name change.*
```{r}
merged <- merged %>%
  mutate(duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")))

sum(is.na(merged$duration))  # 383
```

# Fuzzy Matching 
Rebuild fuzzy matching potential pools
```{r}
# rebuild unmatched pools with clean data
# people who checked in but can't find check out (ie have no timestamp out)
unmatched_in <- data_in %>%
  filter(!id_in %in% (merged %>% filter(!is.na(timestamp_out)) %>% pull(id_in)))

# people who checked out but dont have a matched check in 
unmatched_out <- data_out %>%
  anti_join(merged, by = c("first_name", "last_name", "date", "dob"))

nrow(unmatched_in)   # 383
nrow(unmatched_out)  # 287
```

Fuzzy Matching
First name typo - matches on last name, date and dob
*17 people mistyped first name*
```{r}
tier1_results <- list()

for (i in 1:nrow(unmatched_in)) {
  row_in <- unmatched_in[i, ]
  
  # filter candidates: exact match on last_name, dob, and date
  candidates <- unmatched_out %>%
    filter(
      last_name == row_in$last_name,
      dob       == row_in$dob,
      date      == row_in$date
    )
  
  if (nrow(candidates) == 0) next  # no candidates, skip to next
  
  # fuzzy match on first name only
  dist_first <- stringdist(row_in$first_name, candidates$first_name, method = "jw")
  
  best_idx   <- which.min(dist_first)
  best_score <- dist_first[best_idx]
  
  if (best_score >= 0.3) next  # outside threshold, skip
  
  tier1_results[[i]] <- tibble(
    tier           = 1,
    id_in          = row_in$id_in,
    id_out         = candidates$id_out[best_idx],
    first_name_in  = row_in$first_name,
    last_name_in   = row_in$last_name,
    dob_in         = row_in$dob,
    first_name_out = candidates$first_name[best_idx],
    last_name_out  = candidates$last_name[best_idx],
    dob_out        = candidates$dob[best_idx],
    date           = row_in$date,
    score          = best_score,
    status         = case_when(
      best_score <  0.2 ~ "auto_accept",
      best_score <  0.3 ~ "review",
      TRUE              ~ "no_match"
    )
  )
}

tier1_matches <- bind_rows(tier1_results)

# summary
tier1_matches %>% count(status)

# inspect auto_accept
tier1_matches %>%
  filter(status == "auto_accept") %>%
  select(first_name_in, first_name_out, last_name_in, dob_in, date, score)

# inspect review
tier1_matches %>%
  filter(status == "review") %>%
  select(first_name_in, first_name_out, last_name_in, dob_in, date, score)
tier1_matches %>% count(status)
```

Manually review 2
```{r}
tier1_matches %>%
  filter(status == "review") %>%
  select(id_in, id_out, first_name_in, first_name_out, last_name_in, dob_in, date, score)
```

Accept two manual reviews
```{r}
tier1_matches <- tier1_matches %>%
  mutate(final_status = case_when(
    status == "auto_accept"              ~ "accept",
    status == "review" & id_in %in% c(173, 2542) ~ "accept",
    TRUE                                 ~ "reject"
  ))

tier1_matches %>% count(final_status)
```

Remove accepted first name typos from both pools
```{r}
# remove tier 1 accepted matches from both pools
accepted_in  <- tier1_matches %>% filter(final_status == "accept") %>% pull(id_in)
accepted_out <- tier1_matches %>% filter(final_status == "accept") %>% pull(id_out)

unmatched_in  <- unmatched_in  %>% filter(!id_in  %in% accepted_in)
unmatched_out <- unmatched_out %>% filter(!id_out %in% accepted_out)

nrow(unmatched_in)   # should be 383 - 19 = 364
nrow(unmatched_out)  # should be 287 - 19 = 268
```
DOB Typo
*147 mistyped their birthday*
*Some people mistyped, some people put the date, lots put their birthday but the current year*
```{r}
tier2_results <- list()

for (i in 1:nrow(unmatched_in)) {
  row_in <- unmatched_in[i, ]
  
  # filter candidates: exact match on first_name, last_name, and date
  candidates <- unmatched_out %>%
    filter(
      first_name == row_in$first_name,
      last_name  == row_in$last_name,
      date       == row_in$date
    )
  
  if (nrow(candidates) == 0) next
  
  # fuzzy match on dob only - convert to character so digit typos are caught
  dist_dob <- stringdist(
    as.character(row_in$dob), 
    as.character(candidates$dob), 
    method = "jw"
  )
  
  best_idx   <- which.min(dist_dob)
  best_score <- dist_dob[best_idx]
  
  if (best_score >= 0.3) next
  
  tier2_results[[i]] <- tibble(
    tier           = 2,
    id_in          = row_in$id_in,
    id_out         = candidates$id_out[best_idx],
    first_name_in  = row_in$first_name,
    last_name_in   = row_in$last_name,
    dob_in         = row_in$dob,
    dob_out        = candidates$dob[best_idx],
    first_name_out = candidates$first_name[best_idx],
    last_name_out  = candidates$last_name[best_idx],
    date           = row_in$date,
    score          = best_score,
    status         = case_when(
      best_score <  0.2 ~ "auto_accept",
      best_score <  0.3 ~ "review",
      TRUE              ~ "no_match"
    )
  )
}

tier2_matches <- bind_rows(tier2_results)

tier2_matches %>% count(status)

tier2_matches %>%
  filter(status == "auto_accept") %>%
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, date, score)

tier2_matches %>%
  filter(status == "review") %>%
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, date, score) # 115 accepted, 32 review

# distribution of review cases 
tier2_matches %>%
  filter(status == "review") %>%
  arrange(score) %>%
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, date, score)

tier2_matches %>%
  filter(status == "review") %>%
  summarise(
    min_score  = min(score),
    max_score  = max(score),
    mean_score = mean(score)
  )

# look at the most and least confident review cases
tier2_matches %>%
  filter(status == "review") %>%
  arrange(score) %>%
  slice(c(1:5, (n()-4):n())) %>%  # top 5 and bottom 5
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, score, date)

# accept all review cases
tier2_matches <- tier2_matches %>%
  mutate(final_status = case_when(
    status == "auto_accept" ~ "accept",
    status == "review"      ~ "accept",
    TRUE                    ~ "reject"
  ))

tier2_matches %>% count(final_status) #147 accepted
```

All matched. Remove matched cases from pool
```{r}
accepted_in  <- tier2_matches %>% filter(final_status == "accept") %>% pull(id_in)
accepted_out <- tier2_matches %>% filter(final_status == "accept") %>% pull(id_out)

unmatched_in  <- unmatched_in  %>% filter(!id_in  %in% accepted_in)
unmatched_out <- unmatched_out %>% filter(!id_out %in% accepted_out)

nrow(unmatched_in)   # should be 364 - 147 = 217
nrow(unmatched_out)  # should be 268 - 147 = 121
```

Typos in last name
*17 people mistyped their last name*
```{r}
tier3_results <- list()

for (i in 1:nrow(unmatched_in)) {
  row_in <- unmatched_in[i, ]
  
  # filter candidates: exact match on first_name, dob, and date
  candidates <- unmatched_out %>%
    filter(
      first_name == row_in$first_name,
      dob        == row_in$dob,
      date       == row_in$date
    )
  
  if (nrow(candidates) == 0) next
  
  # fuzzy match on last name only
  dist_last <- stringdist(row_in$last_name, candidates$last_name, method = "jw")
  
  best_idx   <- which.min(dist_last)
  best_score <- dist_last[best_idx]
  
  if (best_score >= 0.3) next
  
  tier3_results[[i]] <- tibble(
    tier           = 3,
    id_in          = row_in$id_in,
    id_out         = candidates$id_out[best_idx],
    first_name_in  = row_in$first_name,
    last_name_in   = row_in$last_name,
    last_name_out  = candidates$last_name[best_idx],
    dob_in         = row_in$dob,
    dob_out        = candidates$dob[best_idx],
    first_name_out = candidates$first_name[best_idx],
    date           = row_in$date,
    score          = best_score,
    status         = case_when(
      best_score <  0.2 ~ "auto_accept",
      best_score <  0.3 ~ "review",
      TRUE              ~ "no_match"
    )
  )
}

tier3_matches <- bind_rows(tier3_results)

tier3_matches %>% count(status)

tier3_matches %>%
  filter(status == "auto_accept") %>%
  select(id_in, id_out, first_name_in, last_name_in, last_name_out, dob_in, date, score)

tier3_matches %>%
  filter(status == "review") %>%
  select(id_in, id_out, first_name_in, last_name_in, last_name_out, dob_in, date, score)

# accept all auto reviews
tier3_matches <- tier3_matches %>%
  mutate(final_status = case_when(
    status == "auto_accept" ~ "accept",
    status == "review"      ~ "accept",
    TRUE                    ~ "reject"
  ))

tier3_matches %>% count(final_status)
```

Remove matched from pool
```{r}
accepted_in  <- tier3_matches %>% filter(final_status == "accept") %>% pull(id_in)
accepted_out <- tier3_matches %>% filter(final_status == "accept") %>% pull(id_out)

unmatched_in  <- unmatched_in  %>% filter(!id_in  %in% accepted_in)
unmatched_out <- unmatched_out %>% filter(!id_out %in% accepted_out)

# 200 people who checked in but cannot find a check out
nrow(unmatched_in)   # should be 217 - 17 = 200
# 104 people who checked out but cannot find a check in
nrow(unmatched_out)  # should be 121 - 17 = 104
```

Determine which instances is right, by looking at frequency table of pair for each type of typo
```{r}
# combine all accepted matches
all_matches <- bind_rows(
  tier1_matches %>% filter(final_status == "accept"),
  tier2_matches %>% filter(final_status == "accept"),
  tier3_matches %>% filter(final_status == "accept")
)

nrow(all_matches)  # should be 183

# build frequency tables from both sides combined for each field
# first name frequency - grouped by last name and dob
first_name_freq <- bind_rows(
  data_in  %>% select(first_name, last_name, dob),
  data_out %>% select(first_name, last_name, dob)
) %>%
  group_by(last_name, dob, first_name) %>%
  summarise(n = n(), .groups = "drop")

# dob frequency - grouped by first name and last name
dob_freq <- bind_rows(
  data_in  %>% select(first_name, last_name, dob),
  data_out %>% select(first_name, last_name, dob)
) %>%
  group_by(first_name, last_name, dob) %>%
  summarise(n = n(), .groups = "drop")

# last name frequency - grouped by first name and dob
last_name_freq <- bind_rows(
  data_in  %>% select(first_name, last_name, dob),
  data_out %>% select(first_name, last_name, dob)
) %>%
  group_by(first_name, dob, last_name) %>%
  summarise(n = n(), .groups = "drop")

# function to pick correct value by frequency, flag ties
pick_by_frequency <- function(val_in, val_out, freq_table, group_vars, field) {
  if (val_in == val_out) return(list(correct = val_in, status = "agree"))
  
  freq_in <- freq_table %>%
    filter(if_all(all_of(group_vars), ~ . == group_vars)) %>%
    filter(.data[[field]] == val_in) %>%
    pull(n) %>%
    sum()
  
  freq_out <- freq_table %>%
    filter(if_all(all_of(group_vars), ~ . == group_vars)) %>%
    filter(.data[[field]] == val_out) %>%
    pull(n) %>%
    sum()
  
  if (freq_in > freq_out)       return(list(correct = val_in,  status = "in_wins"))
  else if (freq_out > freq_in)  return(list(correct = val_out, status = "out_wins"))
  else                          return(list(correct = NA,       status = "tie"))
}

# resolve correct value for each match
corrections <- all_matches %>%
  rowwise() %>%
  mutate(
    correct_value = case_when(
      # tier 1 - resolve first name using last name + dob as group
      tier == 1 ~ {
        freq_in  <- first_name_freq %>%
          filter(last_name == last_name_in, dob == dob_in, first_name == first_name_in) %>%
          pull(n) %>% sum()
        freq_out <- first_name_freq %>%
          filter(last_name == last_name_in, dob == dob_in, first_name == first_name_out) %>%
          pull(n) %>% sum()
        if (first_name_in == first_name_out)   first_name_in
        else if (freq_in >= freq_out)          first_name_in
        else                                   first_name_out
      },
      
      # tier 2 - resolve dob using first name + last name as group
      tier == 2 ~ {
        freq_in  <- dob_freq %>%
          filter(first_name == first_name_in, last_name == last_name_in, dob == dob_in) %>%
          pull(n) %>% sum()
        freq_out <- dob_freq %>%
          filter(first_name == first_name_in, last_name == last_name_in, dob == dob_out) %>%
          pull(n) %>% sum()
        if (as.character(dob_in) == as.character(dob_out))  as.character(dob_in)
        else if (freq_in >= freq_out)                       as.character(dob_in)
        else                                                as.character(dob_out)
      },
      
      # tier 3 - resolve last name using first name + dob as group
      tier == 3 ~ {
        freq_in  <- last_name_freq %>%
          filter(first_name == first_name_in, dob == dob_in, last_name == last_name_in) %>%
          pull(n) %>% sum()
        freq_out <- last_name_freq %>%
          filter(first_name == first_name_in, dob == dob_in, last_name == last_name_out) %>%
          pull(n) %>% sum()
        if (last_name_in == last_name_out)   last_name_in
        else if (freq_in >= freq_out)        last_name_in
        else                                 last_name_out
      }
    ),
    resolution = case_when(
      tier == 1 ~ if_else(first_name_in == first_name_out, "agree",
                  if_else(correct_value == first_name_in,  "in_wins", "out_wins")),
      tier == 2 ~ if_else(as.character(dob_in) == as.character(dob_out), "agree",
                  if_else(correct_value == as.character(dob_in), "in_wins", "out_wins")),
      tier == 3 ~ if_else(last_name_in == last_name_out, "agree",
                  if_else(correct_value == last_name_in,  "in_wins", "out_wins"))
    )
  ) %>%
  ungroup()

# check resolution summary
corrections %>% count(tier, resolution)

# flag any ties for manual review
corrections %>%
  filter(is.na(correct_value)) %>%
  select(tier, id_in, id_out, first_name_in, first_name_out, last_name_in, last_name_out, dob_in, dob_out)
```

From this, we can tell that in typos, it was pretty equal on people mistyping their first and last names at check in and check out. DOB majority of people mistyped at checkout more (check in won at a higher frequency).
*good table for report*

Impute what side is correct 
```{r}
# impute tier 1 corrections - first name
tier1_corrections <- corrections %>% filter(tier == 1)

for (i in 1:nrow(tier1_corrections)) {
  correct <- tier1_corrections$correct_value[i]
  id_in_i  <- tier1_corrections$id_in[i]
  id_out_i <- tier1_corrections$id_out[i]
  
  # fix data_in if it was wrong
  data_in <- data_in %>%
    mutate(first_name = case_when(
      id_in == id_in_i ~ correct,
      TRUE             ~ first_name
    ))
  
  # fix data_out if it was wrong
  data_out <- data_out %>%
    mutate(first_name = case_when(
      id_out == id_out_i ~ correct,
      TRUE               ~ first_name
    ))
}

# impute tier 2 corrections - dob
tier2_corrections <- corrections %>% filter(tier == 2)

for (i in 1:nrow(tier2_corrections)) {
  correct  <- as.Date(tier2_corrections$correct_value[i])
  id_in_i  <- tier2_corrections$id_in[i]
  id_out_i <- tier2_corrections$id_out[i]
  
  data_in <- data_in %>%
    mutate(dob = case_when(
      id_in == id_in_i ~ correct,
      TRUE             ~ dob
    ))
  
  data_out <- data_out %>%
    mutate(dob = case_when(
      id_out == id_out_i ~ correct,
      TRUE               ~ dob
    ))
}

# impute tier 3 corrections - last name
tier3_corrections <- corrections %>% filter(tier == 3)

for (i in 1:nrow(tier3_corrections)) {
  correct  <- tier3_corrections$correct_value[i]
  id_in_i  <- tier3_corrections$id_in[i]
  id_out_i <- tier3_corrections$id_out[i]
  
  data_in <- data_in %>%
    mutate(last_name = case_when(
      id_in == id_in_i ~ correct,
      TRUE             ~ last_name
    ))
  
  data_out <- data_out %>%
    mutate(last_name = case_when(
      id_out == id_out_i ~ correct,
      TRUE               ~ last_name
    ))
}
```

Remerge and check NA
```{r}
merged <- data_in %>%
  left_join(data_out, by = c("date", "dob", "first_name", "last_name"))

merged <- merged %>%
  mutate(
    note = case_when(
      is.na(note.x) & is.na(note.y) ~ NA_character_,
      is.na(note.x)                  ~ note.y,
      is.na(note.y)                  ~ note.x,
      TRUE                           ~ paste(note.x, note.y, sep = " | ")
    )
  ) %>%
  select(-note.x, -note.y)

merged <- merged %>%
  mutate(duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")))

nrow(merged)  # 3788 
nrow(data_in) # 3788
sum(is.na(merged$duration))  # 200
```
3788 people sucessfully checked in and out. 200 people checked in but did not check out (they are in merged with an empty check out). 104 people checked out but never in (they are not in merged because it is a left join)

# IDs
Because DOBs were so messed up, assign based on *distinct first and last name ONLY. This could cause issue because we cannot verify if someone is two different people but with same name except with domain knowledge.*
*When was database collected vs when did checking in start? Did that many people really not come to volunteer or did it not capture soon enough after.*
```{r}
# by first and last name only
merged %>%
  distinct(first_name, last_name) %>%
  nrow() #441

# by first name, last name and dob (stricter - but probably includes mistyped dobs)
merged %>%
  distinct(first_name, last_name, dob) %>%
  nrow() #539

# how many total are in the database?
nrow(database) #2999
nrow(unique(database))
dat

# what date range does your check-in data cover?
range(merged$date) # 1/28/2023 - 2/21/2026 
```

Identify who is in database vs merged
```{r}
# people in database who never appear in merged
database %>%
  anti_join(merged, by = c("first_name", "last_name")) %>%
  nrow() # 2943

# people in merged who are not in database (truly new volunteers)
merged %>%
  distinct(first_name, last_name) %>%
  anti_join(database, by = c("first_name", "last_name")) %>%
  nrow() #388
```


Pull those in database with ID vs not
```{r}
# get unique volunteers from merged - first and last name only
unique_volunteers <- merged %>%
  distinct(first_name, last_name) %>%
  arrange(last_name, first_name)

# standardize database names to match merged formatting
database <- database %>%
  mutate(
    first_name = str_to_lower(str_trim(first_name)),
    last_name  = str_to_lower(str_trim(last_name))
  )

# join to database to pull existing IDs
unique_volunteers <- unique_volunteers %>%
  left_join(
    database %>% select(first_name, last_name, ID),
    by = c("first_name", "last_name")
  )

# check how many matched
sum(!is.na(unique_volunteers$ID))  # found in database
sum(is.na(unique_volunteers$ID))   # need new ID
```

Assign IDs to those not already in database
```{r}
# assign new sequential IDs to those not in database
max_existing_id <- max(database$ID, na.rm = TRUE)

unique_volunteers <- unique_volunteers %>%
  mutate(
    ID = case_when(
      !is.na(ID) ~ ID,
      TRUE        ~ max_existing_id + row_number()
    )
  )

# verify no duplicate IDs
unique_volunteers %>%
  group_by(ID) %>%
  filter(n() > 1)
```

Rejoin all to merged
```{r}
# join IDs back to merged on first + last name only
merged <- merged %>%
  left_join(
    unique_volunteers %>% select(first_name, last_name, ID),
    by = c("first_name", "last_name")
  )

# verify everyone has an ID
sum(is.na(merged$ID))  # should be 0

# verify row count still stable
nrow(merged)  # should still be 3788
nrow(unique(merged$first_name))

# spot check - are names actually comparable?
head(database %>% select(first_name, last_name))
head(unique_volunteers %>% select(first_name, last_name))
```

# flag potential same name cases with  
```{r}
# flag potential same-name different-person cases
unique_volunteers %>%
  group_by(first_name, last_name) %>%
  filter(n() > 1) %>%
  arrange(last_name, first_name)
```

*Recommendation: tell everyone their ID number and have them enter it when signing in; add guardrails to DOB so someone cannot put an age under 18 years old (mitigate people putting todays date). Communicate to use same name - full not nickname or nickname only overall, but especially on same day check in check out. Require email when submitting document - allows robust ground truth robust to mistypes. *