---
title: "volunteer_data"
author: "Caroline Gheen"
date: "2026-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

=============================================================================
 Project: Manos Juntos STATCOM - Volunteer Check In/Out
 Script: volunteer_in_out.R
 Author: Caroline Gheen
 Date: 2026-03-01
 Purpose: Clean and prepare dataset for format in master dataset
=============================================================================

## 1. Packages
## 2. Call in Data and Formatting
## 3. Fix DOB
## 4. Exploring Data
## 5. Merge Datasets with Different Apporoaches (email+date, first+last+date, first+last+dob+date, dob+date)
## 6. Examine Duplicates
## 7. Phone Numbers
## 8. Duration - Total Time Calculation
## 9. Unique ID Assignment

# Install Packages
```{r}
# install packages
# install.packages("readxl")
# install.packages("dplyr")
# install.packages("lubridate", repos = "<https://cloud.r-project.org>")
# install.packages("writexl")
```

# Call Packages
```{r}
library(readxl) 
library(dplyr) 
library(lubridate) 
library(stringr)
library(writexl)
library(fuzzyjoin)
```

# Call in Data and Format
Read in data, telling each column desired structure
```{r}
data_in <- read_excel(
  'Volunteer_Checkin_Data.xlsx', sheet = 'Sign-In',
  col_types = c(
    "date",    # Timestamp_in
    "text",    # Email Address
    "text",    # First Name
    "text",    # Last Name
    "date",    # Date of Birth
    "text",    # Cell Phone Number
    "text",    # Email (non-school...)
    "skip",    # verification question
    "skip",    # covid
    "skip"     # mask
  )
)
nrow(data_in)
```
6 DOBs were outside expected range - default changed these to be NA

Changes names to be in wanted format
```{r}
data_in <- data_in %>%
  rename(
    timestamp_in   = `Timestamp`,
    email       = `Email Address`,
    first_name  = `First Name`,
    last_name   = `Last Name`,
    dob         = `Date of Birth`,
    phone       = `Cell Phone Number`,
    email_personal   = `Email (non-school email please)`
  )
```

Trim and standardize all free text (first/last name, email)
```{r}
data_in <- data_in %>% 
  mutate( 
    first_name = str_to_lower(str_trim(first_name)), 
    last_name = str_to_lower(str_trim(last_name)), 
    email = str_to_lower(str_trim(email)), 
    email_personal = str_to_lower(str_trim(email_personal)) 
)
```

Read in check out data, telling each column desired structure
```{r}
data_out <- read_excel(
  'Volunteer_Checkin_Data.xlsx', sheet = 'Sign-Out',
  col_types = c(
    "date",    # Timestamp
    "text",    # Email Address
    "text",    # First Name
    "text",    # Last Name
    "date",    # Date of Birth
    "text",    # Cell Phone Number
    "skip",    # verification question
    "skip",    # optional
    "text"     # department
  )
)
nrow(data_out)
```
3 DOB were out of range - default changed to NA


Changes names to be in wanted format out data
```{r}
data_out <- data_out %>%
  rename(
    timestamp_out   = `Timestamp`,
    email       = `Email Address`,
    first_name  = `First Name`,
    last_name   = `Last Name`,
    dob         = `Date of Birth`,
    phone       = `Cell Phone Number`,
    department   = `Which department did you spend the majority of your time volunteering in today?`
  )
```

Trim and standardize all free text (first/last name, email)
```{r}
data_out <- data_out %>% 
  mutate(first_name = str_to_lower(str_trim(first_name)), 
  last_name = str_to_lower(str_trim(last_name)), 
  email = str_to_lower(str_trim(email))
)
```

# FIX DOB
Some DOB were misstyped (ex: year is 0200 instead of 2000, assumably). Because there is a low amount, we could go back and manually correct each if needed. But as long as we don't match on dob, we should be fine to leave as is.

Manually imput DOB
```{r}
data_in %>% slice(40, 941, 1298, 1380, 1738, 2306) %>% 
  select(first_name, last_name, dob) # data in mistakes

data_out %>% slice(16, 1280, 1281) %>% 
  select(first_name, last_name, dob) # data out mistakes

# manually fix in dob
data_in <- data_in %>%
  mutate(dob = case_when(
    first_name == "cade" & last_name == "huston" ~ as.Date("1998-09-14"),
    first_name == "paul" & last_name == "atiyeh" ~ as.Date("2004-06-01"),
    first_name == "athman" & last_name == "rajaratnam" ~ as.Date("2001-02-20"),
    first_name == "emily" & last_name == "nguyen" ~ as.Date("2000-09-13"),
    first_name == "adrian" & last_name == "hernandez" ~ as.Date("2003-03-03"),
    first_name == "megan" & last_name == "john" ~ as.Date("2003-10-21"),
    TRUE ~ dob
  ))

# manually fix dob out
data_out <- data_out %>%
  mutate(dob = case_when(
    first_name == "cade" & last_name == "huston" ~ as.Date("1998-09-14"),
    first_name == "emily" & last_name == "nguyen" ~ as.Date("2000-09-13"),
    first_name == "lena" & last_name == "do" ~ as.Date("1998-11-08"),
    TRUE ~ dob
  ))

# check all dob present
sum(is.na(data_in$dob)) #0
sum(is.na(data_out$dob)) #0
```


# EXPLORE DATA 
Explore data (sign-in and out
```{r}
dim(data_in) #3805 and 7 variables
dim(data_out) #3701 and 7 variables - 104 people did not check out
```

Show unique values of each to determine how many different people used check in/out
```{r}
nrow(unique(data_in[c("first_name","last_name")])) #463 people
nrow(unique(data_out[c("first_name","last_name")])) #457 people
```

Need to extract the date out of the timestamps so that they can match - if you try and match on timestamp outright there will be no matches as timestamp is very specific, down to the second
```{r}
data_in <- data_in %>% mutate(date = as.Date(timestamp_in))
data_out <- data_out %>% mutate(date = as.Date(timestamp_out))
```

Is anyone missing time in/time out
```{r}
sum(is.na(data_out$timestamp_out)) # 0
sum(is.na(data_in$timestamp_in)) # 0
```


# MERGING DATASETS TOGETHER 
Merge datasets together
Used left join so that people who checked in but not out, still remained in the merged dataset

Goal is for nrow(data_in) == nrow(merged). using data_in as basis so matching all instances; if there are extra rows, it is due to a many-many issue of the matching not being specific enough and R creating all possible combinations (Ex: john and sara both born on same date volunteer same date, R make 4 possible combinations)
*nrow(data_in) = 3805*
*look at accidental duplication due to system*

I wanted to try two ways to compare approaches, but only 19 people have an email in sign in... is this true??
```{r}
sum(!is.na(data_in$email)) # 19 people do have an email
sum(is.na(data_in$dob))  # 6 people do not have a dob - default
```

Merge based on same email and date
```{r}
merged_email <- data_in %>% 
left_join(data_out, by = c("email", "date"))
```

Merge based on same first, last and date
```{r}
merged_name <- data_in %>% 
left_join(data_out, by = c("first_name","last_name", "date")) 
nrow(merged_name) # 3821
```

Merge based on same dob, first, last and date
*BEST SO FAR*
```{r}
merged_name_dob <- data_in %>% 
left_join(data_out, by = c("first_name", "last_name", "date", "dob")) 
nrow(data_in) # 3805
nrow(merged_name_dob) #3818 -  13 extra, best so far
sum(is.na(merged_name_dob$timestamp_out)) 
anti_join(data_in, data_out, by = c("first_name", "last_name", "date", "dob")) %>% View()

```

Merge based on same dob, and date
```{r}
merged_dob <- data_in %>% 
left_join(data_out, by = c("date", "dob"))
nrow(merged_dob) # 3894
```

# EXAMINE AND CLEAN DUPLICATES 
When we merge based on first and last name, there is a warning that some first+last+date combination appears multiple times. It could be people accidentally filling out the check in when they meant to scan the check out; people doing two shifts in one day; or people with the same name. Most likely the first one, but we shouldn't assume until we confirm

If we look into the duplicate in and outs, we can manually decide and make decisions about where they are supposed to be
*I am doing this one-by-one because we want a robust final product, but we could probably do a loop*
*Would be good to make note of most likely reason for duplication and how many*
*Common Reasons to Have Duplicates on Check In: check in twice; check in instead of out; check in and did not check out*
*Common Reasons to Have Duplicates on Check Out: people thought they were checking out and stayed longer so they checked out twice; checked out instead of in at the beginning; forgot to check in but checked out*

Find the duplicate in data_in
*23 people accidentally checked in instead of out*
```{r}
duplicates_in <- data_in %>% 
group_by(first_name, last_name, date) %>% 
filter(n() > 1) %>% 
arrange(last_name, first_name, date)
```

For loop that filters through duplicates - if they have two check ins and a check out, delete the second check in. If they only have two check ins, move the second check in to check out
```{r}
# get unique duplicate people
duplicates_in <- data_in %>% 
  group_by(first_name, last_name, date, dob) %>% 
  filter(n() > 1) %>% 
  ungroup()

unique_dupes <- duplicates_in %>% 
  distinct(first_name, last_name, date, dob)

for (i in 1:nrow(unique_dupes)) {
  fn   <- unique_dupes$first_name[i]
  ln   <- unique_dupes$last_name[i]
  dt   <- unique_dupes$date[i]
  dob  <- unique_dupes$dob[i]
  
  # check if they have a checkout already
  has_checkout <- nrow(
    data_out %>% filter(first_name == fn, last_name == ln, date == dt)
  ) > 0
  
  if (has_checkout) {
    # they have a checkout already - just delete second check in
    data_in <- data_in %>%
      filter(!(first_name == fn & last_name == ln & 
               date == dt & dob == dob &
               duplicated(cbind(first_name, last_name, date, dob))))
    
  } else {
    # no checkout - move second check in to data_out
    extra_row <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt) %>%
      slice(2) %>%
      rename(timestamp_out = timestamp_in) %>%
      mutate(department = NA_character_) %>%
      select(-email_personal)
    
    data_out <- bind_rows(data_out, extra_row)
    
    # then delete second check in
    data_in <- data_in %>%
      filter(!(first_name == fn & last_name == ln & 
               date == dt &
               duplicated(cbind(first_name, last_name, date, dob))))
  }
}
```
There is one person whose duplicate date is not working - need to manually investigate


Find the duplicate in data_out
```{r}
data_out %>% 
group_by(first_name, last_name, date) %>% 
filter(n() > 1) %>% 
arrange(last_name, first_name, date)
```
One that is not working - need to manually imput

Can create a for loop that deletes the first duplicate entry in data out if there is a data in. If no data in, puts the first entry into data_in
*Before this runs, check how many had duplicate entries for report*

```{r}
duplicates_out <- data_out %>% 
  group_by(first_name, last_name, date, dob) %>% 
  filter(n() > 1) %>% 
  ungroup()

unique_dupes <- duplicates_out %>% 
  distinct(first_name, last_name, date, dob)

for (i in 1:nrow(unique_dupes)) {
  fn  <- unique_dupes$first_name[i]
  ln  <- unique_dupes$last_name[i]
  dt  <- unique_dupes$date[i]
  dob <- unique_dupes$dob[i]
  
  # check if they have a checkin already
  has_checkin <- nrow(
    data_in %>% filter(first_name == fn, last_name == ln, date == dt)
  ) > 0
  
  if (has_checkin) {
    # has a checkin - just delete the duplicate checkout
    data_out <- data_out %>%
      filter(!(first_name == fn & last_name == ln & 
                 date == dt &
                 duplicated(cbind(first_name, last_name, date, dob), fromLast = TRUE)))
    
  } else {
    # no checkin - move first checkout to data_in
    extra_row <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt) %>%
      slice(1) %>%
      rename(timestamp_in = timestamp_out) %>%
      mutate(email_personal = NA_character_) %>%
      select(-department)
    
    data_in <- bind_rows(data_in, extra_row)
    
    # then delete duplicate from data_out
    data_out <- data_out %>%
      filter(!(first_name == fn & last_name == ln & 
               date == dt &
               duplicated(cbind(first_name, last_name, date, dob))))
  }
}

# should return 0 duplicates
data_out %>% 
  group_by(first_name, last_name, date, dob) %>% 
  filter(n() > 1)
```

Remerge after all cleaning
```{r}
merged_name_dob <- data_in %>% 
  left_join(data_out, by = c("first_name", "last_name", "date", "dob"))
```

Check if extra rows from best model: first+last+date+dob
```{r}
merged_name_dob %>% 
group_by(first_name, last_name, date, dob) %>%
filter(n() > 1) %>% 
arrange(last_name, first_name, date) %>% 
View()
```

Check rows; merged == data_in
```{r}
nrow(merged_name_dob) #3785
nrow(data_in) #3785
```
*check how many email personal in checkin match email in checkout*

# DURATION
I want to ensure everyone is lined up correctly before doing duration

Create total time by doing time out - time in
```{r}
merged_name_dob <- merged_name_dob %>% 
  mutate(
  time_in = as.POSIXct(timestamp_in, format = "%Y-%m-%d %H:%M:%S"), 
  time_out = as.POSIXct(timestamp_out, format = "%Y-%m-%d %H:%M:%S"), 
  duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")),
  total_time = sprintf("%dh %dm", as.integer(duration %/% 60),
  as.integer(duration %% 60)) )
View(merged_name_dob)
```

Explore duration data
```{r}
range(na.omit(merged_name_dob$duration)) # -0.8 - 706.305
```

How many are NAs
*205 had mistyped birthdays*
*34 mistyped their first name - some used nicknames, some had additional information, some put last name also*
*__ mistyped last name*
*___ forgot to checkout*

#Relook at dob, there are some people who are putting the date, want to make sure this is not rewriting
```{r}
sum(is.na(merged_name_dob$duration)) # 393 NA - why this many....

# Examine those that are NAs
missing_duration <- merged_name_dob %>% 
  filter(is.na(duration))

# people who have a checkout but failed to match on dob
fixable_dob <- missing_duration %>%
  inner_join(data_out, by = c("first_name", "last_name", "date"))
fixable_name <- missing_duration %>%
  inner_join(data_out, by = c("dob", "last_name", "date"))
nrow(fixable_name) #34 first name
fixable_last <- missing_duration %>%
  inner_join(data_out, by = c("dob", "first_name", "date"))
nrow(fixable_last) #25 last name -129 never checked out 
nrow(data_in) - nrow(data_out) #88 


nrow(fixable_dob) #205 - did not match on dob data_out, fix by imputting dob from check in to their check out.

# create for loop that will match by first, last, date and imput the dob reported at check in to dob at check out
for (i in 1:nrow(fixable_dob)) {
  fn  <- fixable_dob$first_name[i]
  ln  <- fixable_dob$last_name[i]
  dt  <- fixable_dob$date[i]
  correct_dob <- fixable_dob$dob.x[i]  # dob from data_in (check in)
  
  # update dob in data_out to match data_in
  # did i only correct typos in data_out here? what is there is a data_in typo
  
data_out <- data_out %>%
    mutate(dob = case_when(
      first_name == fn & last_name == ln & date == dt ~ correct_dob,
      TRUE ~ dob
    ))
}

# remerge and recreate duration columns 
merged_name_dob <- data_in %>% 
  left_join(data_out, by = c("first_name", "last_name", "date", "dob"))
merged_name_dob <- merged_name_dob %>% 
  mutate(
  time_in = as.POSIXct(timestamp_in, format = "%Y-%m-%d %H:%M:%S"), 
  time_out = as.POSIXct(timestamp_out, format = "%Y-%m-%d %H:%M:%S"), 
  duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")),
  total_time = sprintf("%dh %dm", as.integer(duration %/% 60),
  as.integer(duration %% 60)) )

sum(is.na(merged_name_dob$duration)) # 188

# need to reupdate missing_duration to be the people still missing after fixing dob
missing_duration <- merged_name_dob %>% 
  filter(is.na(duration)) 
nrow(missing_duration) #188
```

People who have a checkout but failed to match on first name
```{r}
# people who have a checkout but failed to match on first name
fixable_first <- missing_duration %>%
  inner_join(data_out, by = c("dob", "last_name", "date"))
nrow(fixable_first) #34

# apply changes to data_out
fixable_first <- fixable_first %>%
  mutate(first_name = case_when(
    first_name == "rgymmmmm"            ~ "amaia",
    first_name == "muhammad.a.ali-1@ou.edu" ~ "muhammad",
    first_name == "4.056578909e9"       ~ "eli",
    TRUE ~ first_name
  ))

# initialize note column; run before the first time adding something to note column
if (!"note" %in% names(fixable_first)) {
  fixable_first <- fixable_first %>% mutate(note = NA_character_)
}

# anything that is after the first space or a /, adds to a note column; if someone has notes in both their check in and check out first name, adds both to note with a | seperator
fixable_first <- fixable_first %>%
  mutate(
    new_note = case_when(
      str_detect(first_name.x, "[ /]") & str_detect(first_name.y, "[ /]") ~ 
        paste(str_extract(first_name.x, "(?<=[ /]).*"), 
              str_extract(first_name.y, "(?<=[ /]).*"), sep = " | "),
      str_detect(first_name.x, "[ /]") ~ str_extract(first_name.x, "(?<=[ /]).*"),
      str_detect(first_name.y, "[ /]") ~ str_extract(first_name.y, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    note = case_when(
      is.na(note) & is.na(new_note) ~ NA_character_,
      is.na(note)                   ~ new_note,
      is.na(new_note)               ~ note,
      TRUE                          ~ paste(note, new_note, sep = " | ")
    ),
    first_name.x = str_trim(str_extract(first_name.x, "^[^ /]+")),
    first_name.y = str_trim(str_extract(first_name.y, "^[^ /]+"))
  ) %>%
  select(-new_note)

# could use fuzzmatch (agrep and adist) to fix
# use data_in as ground truth; find someone who matches on last name and first name and vote for whatever best matches between misspelled check in/check out first name
name_corrections <- data.frame(
  first_name_in  = character(),
  first_name_out = character(),
  correct_name   = character(),
  source         = character(),
  stringsAsFactors = FALSE
)

for (i in 1:nrow(fixable_first)) {
  fn_in  <- fixable_first$first_name.x[i]
  fn_out <- fixable_first$first_name.y[i]
  ln     <- fixable_first$last_name[i]
  dob_i  <- fixable_first$dob[i]
  
  candidate <- data_in %>% 
    filter(last_name == ln, dob == dob_i)
  
  if (nrow(candidate) > 0) {
    db_name <- candidate %>%
      count(first_name) %>% 
      arrange(desc(n)) %>% 
      slice(1) %>% 
      pull(first_name)
  } else {
    db_name <- fn_in  # fall back to check in name if no candidate found
  }
  
  match_x <- agrep(db_name, fn_in,  value = TRUE, max.distance = 0.2)
  match_y <- agrep(db_name, fn_out, value = TRUE, max.distance = 0.2)
  
  correct <- case_when(
    length(match_x) > 0 & length(match_y) > 0 ~ db_name,
    length(match_x) > 0                        ~ fn_in,
    length(match_y) > 0                        ~ fn_out,
    TRUE                                        ~ db_name
  )
  
  name_corrections <- bind_rows(name_corrections, data.frame(
    first_name_in  = fn_in,
    first_name_out = fn_out,
    correct_name   = correct,
    source         = ifelse(nrow(candidate) > 0, "frequency", "no match")
  ))
}

# check what was corrected, to what and from where
name_corrections %>% distinct() %>% arrange(first_name_in)
View(name_corrections)

# imput names that were corrected into fixable first. then imput all of fixable first into data_in and data_out

# imput what is correct into the originally wrong entry
for (i in 1:nrow(name_corrections)) {
  fn_in      <- name_corrections$first_name_in[i]
  fn_out     <- name_corrections$first_name_out[i]
  correct    <- name_corrections$correct_name[i]
  
  # if check in name was wrong, fix it in fixable first
  if (fn_in != correct) {
    fixable_first <- fixable_first %>%
      mutate(first_name.x = case_when(
        first_name.x == fn_in ~ correct,
        TRUE ~ first_name
      ))
  }
  
  # if check out name was wrong, fix it in fixable_first
  if (fn_out != correct) {
    fixable_first <- fixable_first %>%
      mutate(first_name.y = case_when(
        first_name.y == fn_out ~ correct,
        TRUE ~ first_name
      ))
  }
}

# push all of fixable first to data_in and data_out including adding the new notes column
# push corrected first names back to data_in
data_in <- data_in %>%
  left_join(
    fixable_first %>% select(last_name, dob, date, first_name.x, note) %>% 
      rename(first_name = first_name.x),
    by = c("last_name", "dob", "date")
  ) %>%
  mutate(first_name = coalesce(first_name.y, first_name)) %>% mutate(note = note) %>% 
  select(-first_name.y)

# push corrected first names back to data_out
data_out <- data_out %>%
  left_join(
    fixable_first %>% select(last_name, dob, date, first_name.y, note),
    by = c("last_name", "dob", "date")
  ) %>%
  mutate(first_name = coalesce(first_name.y, first_name)) %>% mutate(note = note) %>% 
  select(-first_name.y)

# remerge based on what is fixed
merged_name_dob <- data_in %>% 
  left_join(data_out, by = c("first_name", "last_name", "date", "dob"))
merged_name_dob <- merged_name_dob %>% 
  mutate(
  time_in = as.POSIXct(timestamp_in, format = "%Y-%m-%d %H:%M:%S"), 
  time_out = as.POSIXct(timestamp_out, format = "%Y-%m-%d %H:%M:%S"), 
  duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")),
  total_time = sprintf("%dh %dm", as.integer(duration %/% 60),
  as.integer(duration %% 60)) )
missing_duration <- merged_name_dob %>% 
  filter(is.na(duration))

#ensure missing duration is same as nrow in missing_duration
sum(is.na(merged_name_dob$duration)) #181

# need to reupdate missing_duration to be the people still missing after fixing dob
missing_duration <- merged_name_dob %>% 
  filter(is.na(duration)) 
nrow(missing_duration) #181 - should've fixed 34
```



People with no checkout record at all
```{r}
# people with no checkout record at all - either did not check out or could have mistake in name; create another for loop/do same dob procudure with first and last name to determine spelling issue; any left after those two for loops do not have a check out and will need to have one created 
truly_missing <- missing_duration %>%
  anti_join(data_out, by = c("first_name", "last_name", "date"))
View(truly_missing)
```

Graph duration data
```{r}
hist(na.omit(merged_name_dob$duration))
```


# PHONE NUMBERS 
Phone numbers are coming in as scientific notation
The phone numbers are in the Excel correct, but they are not coming through correctly; we have enough defining features that we should not need the phone numbers, but if we want them to be perfect, we will need to go back and clean them further. Additionally will need to consider if displaying in dashboard. From comparing output to the spreadsheet, I believe it messes up a phone number the same way for all (Ex: Emma S. has a 0 added to the beginning and 1 added to the end of her phone number in all of her entries so her phone number ends up being wrong, but still consistently hers).

Strip all dashes and parentheses to leave only numbers
```{r}
data_in <- data_in %>% 
mutate(phone = str_replace_all(phone, "[\^0-9]", ""))
```

Every phone number should have 10 digits
```{r}
data_in %>% count(nchar(phone)) 
data_in %>% filter(nchar(phone) == 13)
```

Add back in the leading zeros
```{r}
df_in <- df_in %>% 
mutate(phone = as.character(as.numeric(phone))) %>% 
mutate(phone = ifelse(nchar(phone) == 9, paste0("0", phone), phone)
```


# UNIQUE VOLUNTEER ID

Load data
```{r}
database <- read_excel("Volunteer_Database_2024.xlsx", col_types = c(
  "numeric", # ID
  "text", # First Name
  "text", # Last Name
  "text", # Volunteer Address 1
  "text", # Volunteer Address 2
  "text", # City
  "text", # State
  "text", # Zip Code
  "text", # Cell
  "text" # Email
))
```

Changes names to be in wanted format
```{r}
database <- database %>% 
    rename( 
    ID = 'ID', 
    first_name = 'First Name',
    last_name ='Last Name', 
    address1 = 'VolunteerAddress1', 
    address2 ='Volunteer Address2', 
    city = 'VolunteerCity',
    state = 'State', 
    zipcode = 'Zip Code', 
    phone = 'Cell', 
    email = 'Volunteer e-mail',
)
```

Trim and standardize all free text (first/last name, email)
```{r}
database <- database %>% 
  mutate( 
    first_name =str_to_lower(str_trim(first_name)), 
    last_name = str_to_lower(str_trim(last_name)), 
    email = str_to_lower(str_trim(email)),
    address1 = str_to_lower(str_trim(address1)), 
    address2 = str_to_lower(str_trim(address2)), 
    city = str_to_lower(str_trim(city)),
    state = str_to_lower(str_trim(state)) 
)
```

Cycle through ID numbers, drop any duplicates and assign ID numbers to any blanks

Compare to merged dataset and apply next, sequential ID number to those who have not appeared yet

Probably should add leading 0s? look up good practice for id numbers

```{r}
nrow(data_in)
nrow(merged_name_dob)
nrow(unique(merged_name_dob$first_name))
# assign id for the non 393 duplicated
#if else statement to imput corrected name to whatever value is incorrect
```


# Reattempt at fuzzy matching
starting at merged_name_dob
```{r}
nrow(data_in) # 3805
nrow(data_out) # 3701
nrow(data_in) - nrow(data_out) ## 104

merged <- data_in %>% 
left_join(data_out, by = c("date", "dob", "first_name", "last_name"))
nrow(merged_dob) # 3893
```

assign row id individually to data_in and data_out
```{r}
data_in  <- data_in  %>% mutate(id_in  = row_number())
data_out <- data_out %>% mutate(id_out = row_number())
```

Fix duplicate check in - people who submitted two check ins instead of check in check out will have the second check in occurance go to check out
```{r}
# save count for report before modifying
duplicates_in <- data_in %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) %>%
  ungroup()

n_dupes_in <- nrow(duplicates_in)  # save for report

unique_dupes_in <- duplicates_in %>%
  distinct(first_name, last_name, date, dob)

for (i in 1:nrow(unique_dupes_in)) {
  fn    <- unique_dupes_in$first_name[i]
  ln    <- unique_dupes_in$last_name[i]
  dt    <- unique_dupes_in$date[i]
  dob_i <- unique_dupes_in$dob[i]  

  # check if they have a checkout already
  has_checkout <- nrow(
    data_out %>% filter(first_name == fn, last_name == ln, date == dt, dob == dob_i)
  ) > 0

  if (has_checkout) {
    # has a checkout - delete all but first check-in
    rows_to_remove <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%  # keep first, remove the rest
      pull(id_in)

    data_in <- data_in %>% filter(!id_in %in% rows_to_remove)

  } else {
    # no checkout - move all extra check-ins to data_out
    extra_rows <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%  # everything except the first
      rename(timestamp_out = timestamp_in) %>%
      mutate(department = NA_character_) %>%
      select(-email_personal)

    data_out <- bind_rows(data_out, extra_rows)

    # delete the extra check-ins from data_in
    rows_to_remove <- data_in %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%
      pull(id_in)

    data_in <- data_in %>% filter(!id_in %in% rows_to_remove)
  }
}

# verify - should return 0 rows
data_in %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1)
```

Fix duplicate check out - people who submitted two check outs instead of check in check out will have the first check out occurance go to check in
```{r}
# check for any data_out duplicates first (for manual review + report)
duplicates_out <- data_out %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) %>%
  ungroup()

n_dupes_out <- nrow(duplicates_out)  # save for report

# inspect before running loop - flag anything unexpected
duplicates_out %>% arrange(last_name, first_name, date)

unique_dupes_out <- duplicates_out %>%
  distinct(first_name, last_name, date, dob)

for (i in 1:nrow(unique_dupes_out)) {
  fn    <- unique_dupes_out$first_name[i]
  ln    <- unique_dupes_out$last_name[i]
  dt    <- unique_dupes_out$date[i]
  dob_i <- unique_dupes_out$dob[i]

  # check if they have a check-in already
  has_checkin <- nrow(
    data_in %>% filter(first_name == fn, last_name == ln, date == dt, dob == dob_i)
  ) > 0

  if (has_checkin) {
    # has a check-in - delete all but first checkout
    rows_to_remove <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(-1) %>%  # keep first, remove the rest
      pull(id_out)

    data_out <- data_out %>% filter(!id_out %in% rows_to_remove)

  } else {
    # no check-in - move first checkout to data_in
    extra_row <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      slice(1) %>%
      rename(timestamp_in = timestamp_out) %>%
      mutate(email_personal = NA_character_) %>%
      select(-department)

    data_in <- bind_rows(data_in, extra_row)

    # delete all duplicates from data_out
    rows_to_remove <- data_out %>%
      filter(first_name == fn, last_name == ln, date == dt, dob == dob_i) %>%
      pull(id_out)

    data_out <- data_out %>% filter(!id_out %in% rows_to_remove)
  }
}

# verify - should return 0 rows
data_out %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1)
```

```{r}
# 1. did row counts change as expected?
nrow(data_in)# 3788
nrow(data_out) #3692

# 2. are there any remaining duplicates in data_in?
data_in %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) #0

# 3. are there any remaining duplicates in data_out?
data_out %>%
  group_by(first_name, last_name, date, dob) %>%
  filter(n() > 1) #0

# remerge
merged <- data_in %>% 
left_join(data_out, by = c("date", "dob", "first_name", "last_name"))
nrow(merged) # 3788
nrow(data_in) #3788
```

Ok, now there are no extra check in occurances. Make duration column and check how many NA
```{r}
# calculate duration
merged <- merged %>%
  mutate(duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")))

# then check NAs
sum(is.na(merged$duration)) #398 NA

# and see what they look like
merged %>%
  filter(is.na(duration)) %>%
  select(first_name, last_name, date, dob, timestamp_in, timestamp_out) %>% View()
```
Clean id columns from rbind
```{r}
# remove artifact columns from data_in and data_out
data_in  <- data_in  %>% select(-any_of("id_out"))
data_out <- data_out %>% select(-any_of("id_in"))

# reassign clean IDs
data_in  <- data_in  %>% mutate(id_in  = row_number())
data_out <- data_out %>% mutate(id_out = row_number())

# remerge cleanly
merged <- data_in %>%
  left_join(data_out, by = c("date", "dob", "first_name", "last_name"))

names(merged)  # should now be clean with no .x .y on IDs
```

Pull in all unmatched checkouts
```{r}
# these are unmatched AND clean enough to have joined correctly
unmatched_out_clean <- data_out %>%
  anti_join(merged, by = c("first_name", "last_name", "date", "dob"))
nrow(unmatched_out_clean) #302

# but the TRUE unmatched pool is actually the NAs in merged
truly_unmatched_in <- merged %>%
  filter(is.na(timestamp_out))
nrow(truly_unmatched_in) #398
```

Clean data in of extra wording in either first name in or out and append to note column 
```{r}
# initialize note column in data_in
if (!"note" %in% names(data_in)) {
  data_in <- data_in %>% mutate(note = NA_character_)
}

data_in <- data_in %>%
  mutate(
    new_note_first = case_when(
      str_detect(first_name, "[ /]") ~ str_extract(first_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note_last = case_when(
      str_detect(last_name, "[ /]") ~ str_extract(last_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note = case_when(
      !is.na(new_note_first) & !is.na(new_note_last) ~ paste(new_note_first, new_note_last, sep = " | "),
      !is.na(new_note_first) ~ new_note_first,
      !is.na(new_note_last)  ~ new_note_last,
      TRUE ~ NA_character_
    ),
    note = case_when(
      is.na(note) & is.na(new_note) ~ NA_character_,
      is.na(note)                   ~ new_note,
      is.na(new_note)               ~ note,
      TRUE                          ~ paste(note, new_note, sep = " | ")
    ),
    first_name = str_trim(str_extract(first_name, "^[^ /]+")),
    last_name  = str_trim(str_extract(last_name,  "^[^ /]+"))
  ) %>%
  select(-new_note_first, -new_note_last, -new_note)
```

Clean data out of extra wording in either first name in or out and append to note column 
```{r}
# initialize note column in data_out
if (!"note" %in% names(data_out)) {
  data_out <- data_out %>% mutate(note = NA_character_)
}

data_out <- data_out %>%
  mutate(
    new_note_first = case_when(
      str_detect(first_name, "[ /]") ~ str_extract(first_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note_last = case_when(
      str_detect(last_name, "[ /]") ~ str_extract(last_name, "(?<=[ /]).*"),
      TRUE ~ NA_character_
    ),
    new_note = case_when(
      !is.na(new_note_first) & !is.na(new_note_last) ~ paste(new_note_first, new_note_last, sep = " | "),
      !is.na(new_note_first) ~ new_note_first,
      !is.na(new_note_last)  ~ new_note_last,
      TRUE ~ NA_character_
    ),
    note = case_when(
      is.na(note) & is.na(new_note) ~ NA_character_,
      is.na(note)                   ~ new_note,
      is.na(new_note)               ~ note,
      TRUE                          ~ paste(note, new_note, sep = " | ")
    ),
    first_name = str_trim(str_extract(first_name, "^[^ /]+")),
    last_name  = str_trim(str_extract(last_name,  "^[^ /]+"))
  ) %>%
  select(-new_note_first, -new_note_last, -new_note)
```

Remerge and reconcile notes 
```{r}
# remerge with clean names
merged <- data_in %>%
  left_join(data_out, by = c("date", "dob", "first_name", "last_name"))

# reconcile notes from both sides into a single note column
merged <- merged %>%
  mutate(
    note = case_when(
      is.na(note.x) & is.na(note.y) ~ NA_character_,
      is.na(note.x)                  ~ note.y,
      is.na(note.y)                  ~ note.x,
      TRUE                           ~ paste(note.x, note.y, sep = " | ")
    )
  ) %>%
  select(-note.x, -note.y)

# verify row count still matches
nrow(merged)   # should equal nrow(data_in) = 3788

# check how notes look
merged %>%
  filter(!is.na(note)) %>%
  select(first_name, last_name, date, note)
```

Recalculate duration and check NA
```{r}
merged <- merged %>%
  mutate(duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")))

sum(is.na(merged$duration))  # 383
```

Rebuild fuzzy matching potential pools
```{r}
# rebuild unmatched pools with clean data
# people who checked in but can't find check out (ie have no timestamp out)
unmatched_in <- data_in %>%
  filter(!id_in %in% (merged %>% filter(!is.na(timestamp_out)) %>% pull(id_in)))

# people who checked out but dont have a matched check in 
unmatched_out <- data_out %>%
  anti_join(merged, by = c("first_name", "last_name", "date", "dob"))

nrow(unmatched_in)   # 383
nrow(unmatched_out)  # 287
```

Fuzzy Matching
First name typo - matches on last name, date and dob
*17 people mistyped first name*
```{r}
library(stringdist)

tier1_results <- list()

for (i in 1:nrow(unmatched_in)) {
  row_in <- unmatched_in[i, ]
  
  # filter candidates: exact match on last_name, dob, and date
  candidates <- unmatched_out %>%
    filter(
      last_name == row_in$last_name,
      dob       == row_in$dob,
      date      == row_in$date
    )
  
  if (nrow(candidates) == 0) next  # no candidates, skip to next
  
  # fuzzy match on first name only
  dist_first <- stringdist(row_in$first_name, candidates$first_name, method = "jw")
  
  best_idx   <- which.min(dist_first)
  best_score <- dist_first[best_idx]
  
  if (best_score >= 0.3) next  # outside threshold, skip
  
  tier1_results[[i]] <- tibble(
    tier           = 1,
    id_in          = row_in$id_in,
    id_out         = candidates$id_out[best_idx],
    first_name_in  = row_in$first_name,
    last_name_in   = row_in$last_name,
    dob_in         = row_in$dob,
    first_name_out = candidates$first_name[best_idx],
    last_name_out  = candidates$last_name[best_idx],
    dob_out        = candidates$dob[best_idx],
    date           = row_in$date,
    score          = best_score,
    status         = case_when(
      best_score <  0.2 ~ "auto_accept",
      best_score <  0.3 ~ "review",
      TRUE              ~ "no_match"
    )
  )
}

tier1_matches <- bind_rows(tier1_results)

# summary
tier1_matches %>% count(status)

# inspect auto_accept
tier1_matches %>%
  filter(status == "auto_accept") %>%
  select(first_name_in, first_name_out, last_name_in, dob_in, date, score)

# inspect review
tier1_matches %>%
  filter(status == "review") %>%
  select(first_name_in, first_name_out, last_name_in, dob_in, date, score)
tier1_matches %>% count(status)
```

Manually review 2
```{r}
tier1_matches %>%
  filter(status == "review") %>%
  select(id_in, id_out, first_name_in, first_name_out, last_name_in, dob_in, date, score)
```

Accept two manual reviews
```{r}
tier1_matches <- tier1_matches %>%
  mutate(final_status = case_when(
    status == "auto_accept"              ~ "accept",
    status == "review" & id_in %in% c(173, 2542) ~ "accept",
    TRUE                                 ~ "reject"
  ))

tier1_matches %>% count(final_status)
```
Remove accepted first name typos from both pools
```{r}
# remove tier 1 accepted matches from both pools
accepted_in  <- tier1_matches %>% filter(final_status == "accept") %>% pull(id_in)
accepted_out <- tier1_matches %>% filter(final_status == "accept") %>% pull(id_out)

unmatched_in  <- unmatched_in  %>% filter(!id_in  %in% accepted_in)
unmatched_out <- unmatched_out %>% filter(!id_out %in% accepted_out)

nrow(unmatched_in)   # should be 383 - 19 = 364
nrow(unmatched_out)  # should be 287 - 19 = 268
```
DOB Typo
*147 mistyped their birthday*
*Some people mistyped, some people put the date, lots put their birthday but the current year*
```{r}
tier2_results <- list()

for (i in 1:nrow(unmatched_in)) {
  row_in <- unmatched_in[i, ]
  
  # filter candidates: exact match on first_name, last_name, and date
  candidates <- unmatched_out %>%
    filter(
      first_name == row_in$first_name,
      last_name  == row_in$last_name,
      date       == row_in$date
    )
  
  if (nrow(candidates) == 0) next
  
  # fuzzy match on dob only - convert to character so digit typos are caught
  dist_dob <- stringdist(
    as.character(row_in$dob), 
    as.character(candidates$dob), 
    method = "jw"
  )
  
  best_idx   <- which.min(dist_dob)
  best_score <- dist_dob[best_idx]
  
  if (best_score >= 0.3) next
  
  tier2_results[[i]] <- tibble(
    tier           = 2,
    id_in          = row_in$id_in,
    id_out         = candidates$id_out[best_idx],
    first_name_in  = row_in$first_name,
    last_name_in   = row_in$last_name,
    dob_in         = row_in$dob,
    dob_out        = candidates$dob[best_idx],
    first_name_out = candidates$first_name[best_idx],
    last_name_out  = candidates$last_name[best_idx],
    date           = row_in$date,
    score          = best_score,
    status         = case_when(
      best_score <  0.2 ~ "auto_accept",
      best_score <  0.3 ~ "review",
      TRUE              ~ "no_match"
    )
  )
}

tier2_matches <- bind_rows(tier2_results)

tier2_matches %>% count(status)

tier2_matches %>%
  filter(status == "auto_accept") %>%
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, date, score)

tier2_matches %>%
  filter(status == "review") %>%
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, date, score) # 115 accepted, 32 review

# distribution of review cases 
tier2_matches %>%
  filter(status == "review") %>%
  arrange(score) %>%
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, date, score)

tier2_matches %>%
  filter(status == "review") %>%
  summarise(
    min_score  = min(score),
    max_score  = max(score),
    mean_score = mean(score)
  )

# look at the most and least confident review cases
tier2_matches %>%
  filter(status == "review") %>%
  arrange(score) %>%
  slice(c(1:5, (n()-4):n())) %>%  # top 5 and bottom 5
  select(id_in, id_out, first_name_in, last_name_in, dob_in, dob_out, score, date)

# accept all review cases
tier2_matches <- tier2_matches %>%
  mutate(final_status = case_when(
    status == "auto_accept" ~ "accept",
    status == "review"      ~ "accept",
    TRUE                    ~ "reject"
  ))

tier2_matches %>% count(final_status) #147 accepted
```

Remove matched cases from pool
```{r}
accepted_in  <- tier2_matches %>% filter(final_status == "accept") %>% pull(id_in)
accepted_out <- tier2_matches %>% filter(final_status == "accept") %>% pull(id_out)

unmatched_in  <- unmatched_in  %>% filter(!id_in  %in% accepted_in)
unmatched_out <- unmatched_out %>% filter(!id_out %in% accepted_out)

nrow(unmatched_in)   # should be 364 - 147 = 217
nrow(unmatched_out)  # should be 268 - 147 = 121
```

Typos in last name
**
```{r}
tier3_results <- list()

for (i in 1:nrow(unmatched_in)) {
  row_in <- unmatched_in[i, ]
  
  # filter candidates: exact match on first_name, dob, and date
  candidates <- unmatched_out %>%
    filter(
      first_name == row_in$first_name,
      dob        == row_in$dob,
      date       == row_in$date
    )
  
  if (nrow(candidates) == 0) next
  
  # fuzzy match on last name only
  dist_last <- stringdist(row_in$last_name, candidates$last_name, method = "jw")
  
  best_idx   <- which.min(dist_last)
  best_score <- dist_last[best_idx]
  
  if (best_score >= 0.3) next
  
  tier3_results[[i]] <- tibble(
    tier           = 3,
    id_in          = row_in$id_in,
    id_out         = candidates$id_out[best_idx],
    first_name_in  = row_in$first_name,
    last_name_in   = row_in$last_name,
    last_name_out  = candidates$last_name[best_idx],
    dob_in         = row_in$dob,
    dob_out        = candidates$dob[best_idx],
    first_name_out = candidates$first_name[best_idx],
    date           = row_in$date,
    score          = best_score,
    status         = case_when(
      best_score <  0.2 ~ "auto_accept",
      best_score <  0.3 ~ "review",
      TRUE              ~ "no_match"
    )
  )
}

tier3_matches <- bind_rows(tier3_results)

tier3_matches %>% count(status)

tier3_matches %>%
  filter(status == "auto_accept") %>%
  select(id_in, id_out, first_name_in, last_name_in, last_name_out, dob_in, date, score)

tier3_matches %>%
  filter(status == "review") %>%
  select(id_in, id_out, first_name_in, last_name_in, last_name_out, dob_in, date, score)

# accept all auto reviews
tier3_matches <- tier3_matches %>%
  mutate(final_status = case_when(
    status == "auto_accept" ~ "accept",
    status == "review"      ~ "accept",
    TRUE                    ~ "reject"
  ))

tier3_matches %>% count(final_status)
```

Remove matched from pool
```{r}
accepted_in  <- tier3_matches %>% filter(final_status == "accept") %>% pull(id_in)
accepted_out <- tier3_matches %>% filter(final_status == "accept") %>% pull(id_out)

unmatched_in  <- unmatched_in  %>% filter(!id_in  %in% accepted_in)
unmatched_out <- unmatched_out %>% filter(!id_out %in% accepted_out)

# 200 people who checked in but cannot find a check out
nrow(unmatched_in)   # should be 217 - 17 = 200
# 104 people who checked out but cannot find a check in
nrow(unmatched_out)  # should be 121 - 17 = 104
```

Determine which instances is right, by looking at frequency table of pair for each type of typo
```{r}
# combine all accepted matches
all_matches <- bind_rows(
  tier1_matches %>% filter(final_status == "accept"),
  tier2_matches %>% filter(final_status == "accept"),
  tier3_matches %>% filter(final_status == "accept")
)

nrow(all_matches)  # should be 183

# build frequency tables from both sides combined for each field
# first name frequency - grouped by last name and dob
first_name_freq <- bind_rows(
  data_in  %>% select(first_name, last_name, dob),
  data_out %>% select(first_name, last_name, dob)
) %>%
  group_by(last_name, dob, first_name) %>%
  summarise(n = n(), .groups = "drop")

# dob frequency - grouped by first name and last name
dob_freq <- bind_rows(
  data_in  %>% select(first_name, last_name, dob),
  data_out %>% select(first_name, last_name, dob)
) %>%
  group_by(first_name, last_name, dob) %>%
  summarise(n = n(), .groups = "drop")

# last name frequency - grouped by first name and dob
last_name_freq <- bind_rows(
  data_in  %>% select(first_name, last_name, dob),
  data_out %>% select(first_name, last_name, dob)
) %>%
  group_by(first_name, dob, last_name) %>%
  summarise(n = n(), .groups = "drop")

# function to pick correct value by frequency, flag ties
pick_by_frequency <- function(val_in, val_out, freq_table, group_vars, field) {
  if (val_in == val_out) return(list(correct = val_in, status = "agree"))
  
  freq_in <- freq_table %>%
    filter(if_all(all_of(group_vars), ~ . == group_vars)) %>%
    filter(.data[[field]] == val_in) %>%
    pull(n) %>%
    sum()
  
  freq_out <- freq_table %>%
    filter(if_all(all_of(group_vars), ~ . == group_vars)) %>%
    filter(.data[[field]] == val_out) %>%
    pull(n) %>%
    sum()
  
  if (freq_in > freq_out)       return(list(correct = val_in,  status = "in_wins"))
  else if (freq_out > freq_in)  return(list(correct = val_out, status = "out_wins"))
  else                          return(list(correct = NA,       status = "tie"))
}

# resolve correct value for each match
corrections <- all_matches %>%
  rowwise() %>%
  mutate(
    correct_value = case_when(
      # tier 1 - resolve first name using last name + dob as group
      tier == 1 ~ {
        freq_in  <- first_name_freq %>%
          filter(last_name == last_name_in, dob == dob_in, first_name == first_name_in) %>%
          pull(n) %>% sum()
        freq_out <- first_name_freq %>%
          filter(last_name == last_name_in, dob == dob_in, first_name == first_name_out) %>%
          pull(n) %>% sum()
        if (first_name_in == first_name_out)   first_name_in
        else if (freq_in >= freq_out)          first_name_in
        else                                   first_name_out
      },
      
      # tier 2 - resolve dob using first name + last name as group
      tier == 2 ~ {
        freq_in  <- dob_freq %>%
          filter(first_name == first_name_in, last_name == last_name_in, dob == dob_in) %>%
          pull(n) %>% sum()
        freq_out <- dob_freq %>%
          filter(first_name == first_name_in, last_name == last_name_in, dob == dob_out) %>%
          pull(n) %>% sum()
        if (as.character(dob_in) == as.character(dob_out))  as.character(dob_in)
        else if (freq_in >= freq_out)                       as.character(dob_in)
        else                                                as.character(dob_out)
      },
      
      # tier 3 - resolve last name using first name + dob as group
      tier == 3 ~ {
        freq_in  <- last_name_freq %>%
          filter(first_name == first_name_in, dob == dob_in, last_name == last_name_in) %>%
          pull(n) %>% sum()
        freq_out <- last_name_freq %>%
          filter(first_name == first_name_in, dob == dob_in, last_name == last_name_out) %>%
          pull(n) %>% sum()
        if (last_name_in == last_name_out)   last_name_in
        else if (freq_in >= freq_out)        last_name_in
        else                                 last_name_out
      }
    ),
    resolution = case_when(
      tier == 1 ~ if_else(first_name_in == first_name_out, "agree",
                  if_else(correct_value == first_name_in,  "in_wins", "out_wins")),
      tier == 2 ~ if_else(as.character(dob_in) == as.character(dob_out), "agree",
                  if_else(correct_value == as.character(dob_in), "in_wins", "out_wins")),
      tier == 3 ~ if_else(last_name_in == last_name_out, "agree",
                  if_else(correct_value == last_name_in,  "in_wins", "out_wins"))
    )
  ) %>%
  ungroup()

# check resolution summary
corrections %>% count(tier, resolution)

# flag any ties for manual review
corrections %>%
  filter(is.na(correct_value)) %>%
  select(tier, id_in, id_out, first_name_in, first_name_out, last_name_in, last_name_out, dob_in, dob_out)
```

From this, we can tell that in typos, it was pretty equal on people mistyping their first and last names at check in and check out. DOB majority of people mistyped at checkout more (check in won at a higher frequency).
*good table for report*

Impute what side is correct 
```{r}
# impute tier 1 corrections - first name
tier1_corrections <- corrections %>% filter(tier == 1)

for (i in 1:nrow(tier1_corrections)) {
  correct <- tier1_corrections$correct_value[i]
  id_in_i  <- tier1_corrections$id_in[i]
  id_out_i <- tier1_corrections$id_out[i]
  
  # fix data_in if it was wrong
  data_in <- data_in %>%
    mutate(first_name = case_when(
      id_in == id_in_i ~ correct,
      TRUE             ~ first_name
    ))
  
  # fix data_out if it was wrong
  data_out <- data_out %>%
    mutate(first_name = case_when(
      id_out == id_out_i ~ correct,
      TRUE               ~ first_name
    ))
}

# impute tier 2 corrections - dob
tier2_corrections <- corrections %>% filter(tier == 2)

for (i in 1:nrow(tier2_corrections)) {
  correct  <- as.Date(tier2_corrections$correct_value[i])
  id_in_i  <- tier2_corrections$id_in[i]
  id_out_i <- tier2_corrections$id_out[i]
  
  data_in <- data_in %>%
    mutate(dob = case_when(
      id_in == id_in_i ~ correct,
      TRUE             ~ dob
    ))
  
  data_out <- data_out %>%
    mutate(dob = case_when(
      id_out == id_out_i ~ correct,
      TRUE               ~ dob
    ))
}

# impute tier 3 corrections - last name
tier3_corrections <- corrections %>% filter(tier == 3)

for (i in 1:nrow(tier3_corrections)) {
  correct  <- tier3_corrections$correct_value[i]
  id_in_i  <- tier3_corrections$id_in[i]
  id_out_i <- tier3_corrections$id_out[i]
  
  data_in <- data_in %>%
    mutate(last_name = case_when(
      id_in == id_in_i ~ correct,
      TRUE             ~ last_name
    ))
  
  data_out <- data_out %>%
    mutate(last_name = case_when(
      id_out == id_out_i ~ correct,
      TRUE               ~ last_name
    ))
}
```

Remerge and check NA
```{r}
merged <- data_in %>%
  left_join(data_out, by = c("date", "dob", "first_name", "last_name"))

merged <- merged %>%
  mutate(
    note = case_when(
      is.na(note.x) & is.na(note.y) ~ NA_character_,
      is.na(note.x)                  ~ note.y,
      is.na(note.y)                  ~ note.x,
      TRUE                           ~ paste(note.x, note.y, sep = " | ")
    )
  ) %>%
  select(-note.x, -note.y)

merged <- merged %>%
  mutate(duration = as.numeric(difftime(timestamp_out, timestamp_in, units = "mins")))

nrow(merged)  # 3788 
nrow(data_in) # 3788
sum(is.na(merged$duration))  # 200
```
3788 people sucessfully checked in and out. 200 people checked in but did not check out (they are in merged with an empty check out). 104 people checked out but never in (they are not in merged because it is a left join)

# IDs

Assign new IDs to volunteers not in database and apply existing ID to volunteer records
```{r}
# get unique volunteers from merged
unique_volunteers <- merged %>%
  distinct(first_name, last_name, dob) %>%
  arrange(last_name, first_name)

# standardize database names to match merged formatting
database <- database %>%
  mutate(
    first_name = str_to_lower(str_trim(first_name)),
    last_name  = str_to_lower(str_trim(last_name))
  )

# join to database to pull existing IDs
unique_volunteers <- unique_volunteers %>%
  left_join(
    database %>% select(first_name, last_name, ID),
    by = c("first_name", "last_name")
  )

# check how many matched
sum(!is.na(unique_volunteers$ID))  # found in database - 543
sum(is.na(unique_volunteers$ID))   # need new ID

# assign new sequential IDs to those not in database
max_existing_id <- max(database$ID, na.rm = TRUE)

unique_volunteers <- unique_volunteers %>%
  mutate(
    ID = case_when(
      !is.na(ID) ~ ID,  # keep existing ID
      TRUE        ~ max_existing_id + row_number()  # assign new sequential ID
    )
  )

# verify no duplicate IDs
unique_volunteers %>%
  group_by(ID) %>%
  filter(n() > 1)

# join IDs back to merged
merged <- merged %>%
  left_join(
    unique_volunteers %>% select(first_name, last_name, dob, ID),
    by = c("first_name", "last_name", "dob")
  )

# verify everyone has an ID
sum(is.na(merged$ID))  # should be 0
```


```{r}
# flag potential same-name different-person cases
unique_volunteers %>%
  group_by(first_name, last_name) %>%
  filter(n() > 1) %>%
  arrange(last_name, first_name)
```

